Using high-resolution variant frequencies to empower clinical genome interpretation

### 2016-10-03 manuscript/manuscript.Rmd
Removed more words

In order to pre-compute `af_filter` values for all of ExAC (verson 0.3.1), we apply a two-step approach to the AC and AN values for each of the five major continental populations, and take the highest result from any population.

@@

More explanation can be found in **Supplementary Methods**.

@@

After curating variants above our calculated HCM threshold, the false discovery rate was 0/477 (0.000; 95%CI 0.000-0.008) and 1/559 (0.002; 95%CI 0.000-0.010) for the published HCM cohort and ClinVar data respectively.

@@

In order to assist others in applying our framework, we have precomputed a 'filtering allele frequency' for all variants across the ExAC dataset. This is defined such that if the filtering allele frequency of a variant is at or above the user-defined "maximum credible population allele frequency" for the disease in question, then that variant is not a credible candidate (in other words, for any population allele frequency below the threshold value, the probability of the observed allele count in the ExAC sample is <0.05). Once a user has determined their "maximum credible population allele frequency", they may remove from consideration ExAC variants for which the filtering allele frequency is greater than or equal to than the chosen value.

@@

We recognize several limitations of our approach. First, the approach is limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches to estimation and extrapolation of parameters. In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations. It is critical to define the genetic architecture in the population under study. We acknowledge that there is no substitute for knowning the genetic architecture of the condition of interest, and for some conditions this is not possible. It is the nature of any framework to rely on underlying assumptions, and we believe our method makes these explicit. If a parameter is particularly unclear, it is possible to define a range of architectures that fit with the observed variant. Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants (see **Supplementary Methods** for alternative approaches).

Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some Mendelian conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts. Where this is possible, the maximum credible population allele frequency can be simply computed based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence. Finally, although the resulting allele frequency thresholds are more stringent than those previously used, they are likely to still be very lenient for many applications. For instance, we base our calculation on the most prevalent known pathogenic variant from a disease cohort. For HCM, for which more than 6,000 people have been sequenced, it is unlikely that any single newly identified variant, not previously catalogued in this large cohort, will explain a similarly large proportion of the disease as the most common causal variant, at least in well-studied populations. Future work may therefore involve modeling the frequency distribution of all known variants for a disorder, to further refine these thresholds.

The power of our approach is limited by currently available datasets. Increases in both the ancestral diversity and size of reference datasets will bring additional power to our method over time. We have avoided filtering on variants observed only once, because a single observation provides little information about true allele frequency (**Supplementary Methods**). A ten-fold increase in sample size, resulting from projects such as the US Precision Medicine Initiative, will separate vanishingly rare variants from those whose frequency really is ~1 in 100,000. Increased phenotypic information linked to reference datasets will also reduce limitations due to uncertain disease status, and improve prevalence estimates, adding further power to our approach.

>>>>

In order to pre-compute `af_filter` values for all of ExAC (verson 0.3.1), we apply a two-step approach to the AC and AN values for each of the five major continental populations, and take the highest result from any population (more explanation in **Supplementary Methods**).

@@

@@

@@

To assist others in applying our framework, we have precomputed a 'filtering allele frequency' for all variants across the ExAC dataset. This is defined such that if the filtering allele frequency of a variant is at or above the user-defined "maximum credible population allele frequency" for the disease in question, then that variant is not a credible candidate (in other words, for any population allele frequency below the threshold value, the probability of the observed allele count in the ExAC sample is <0.05). Once a user has determined their "maximum credible population allele frequency", they may remove from consideration ExAC variants for which the filtering allele frequency is greater than or equal to than the chosen value.

@@

We recognise several limitations of our approach. First, the approach is limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters. In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations. It is critical to define the genetic architecture in the population under study. We acknowledge that there is no substitute for knowning the genetic architecture of the condition of interest, and for some conditions this is not possible. It is the nature of any framework to rely on underlying assumptions, and we believe our method makes these explicit. If a parameter is particularly unclear, it is possible to define a range of architectures that fit with the observed variant. Secondly, it is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume for newly discovered pathogenic variants (see **Supplementary Methods** for alternative approaches).

Thirdly, while we believe that ExAC is depleted of severe childhood inherited conditions, and not enriched for cardiomyopathies, it could be enriched relative to the general population for some conditions, including Mendelian forms of common diseases such as diabetes or coronary disease that have been studied in contributing cohorts. Where this is possible, the maximum credible population allele frequency can be simply computed based on the estimated disease prevalence in the ExAC cohort, rather than the population prevalence. Finally, although the resulting allele frequency thresholds are more stringent than those previously used, they are likely to still be very lenient for many applications. For instance, we base our calculation on the most prevalent known pathogenic variant from a disease cohort. For HCM, for which more than 6,000 people have been sequenced, it is unlikely that any single newly identified variant, not previously catalogued in this large cohort, will explain a similarly large proportion of the disease as the most common causal variant, at least in well-studied populations. Future work may therefore involve modeling the frequency distribution of all known variants for a disorder, to further refine these thresholds.

The power of our approach is limited by currently available datasets. Increases in both the ancestral diversity and size of reference datasets will bring additional power to our method over time. We have avoided filtering on variants observed only once, because a single observation provides little information about true allele frequency (see **Supplementary Methods**). A ten-fold increase in sample size, resulting from projects such as the US Precision Medicine Initiative, will separate vanishingly rare variants from those whose frequency really is ~1 in 100,000. Increased phenotypic information linked to reference datasets will also reduce limitations due to uncertain disease status, and improve prevalence estimates, adding further power to our approach.
